---
title: "Introducing read_waterdata_statistics_*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing read_waterdata_statistics_*}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `read_waterdata_stats_por` and `read_waterdata_stats_daterange` functions replace the legacy `readNWISstat` function.
This replacement is necessary because the legacy API service that `readNWISstat` will be decommissioned and replaced with a [modernized API](https://api.waterdata.usgs.gov/statistics/v0/docs).
This new API has two available endpoints, `observationNormals` and `observationIntervals`, that appear similar at first yet have important differences we want to highlight here.

```{r setup}
library(dataRetrieval)

site1 <- "USGS-05428500"
```

## Fetching day-of-year and month-of-year statistics

Day-of-year and month-of-year statistics aggregate observations for the same calendar day or month across multiple years to describe typical seasonal conditions (e.g., all Januarys or all January 1sts).
Consider the output below, where we request day-of-year discharge averages for January 1 and January 2.
Note that the `start_date` and `end_date` are set in `month-year` format to describe the day-of-year range.

```{r, message=FALSE, warning=FALSE}
jan_por_mean <-
  read_waterdata_stats_por(
  monitoring_location_id = site1,
  parameter_code = "00060",
  computation = "arithmetic_mean",
  start_date = "01-01",
  end_date = "01-02"
)

jan_por_mean
```

The first two rows show the average discharge values, aggregated across all January 1sts and 2nds in the site's Period of Record (POR).
But wait: what's in that third row?
Looking at the `time_of_year` and `time_of_year_type` columns, we see the third row represents the average discharge value aggregated across all *Januarys*.
This illustrates one quirk of the modern statistics API: any time the `start_date` to `end_date` range overlaps with the first day of a month (e.g., `"01-01"`), we will get month-of-year as well as the day-of-year summary statistics.
You can filter these rows out of the data if you don't want them in downstream analyses:

```{r}
jan_por_mean[jan_por_mean$time_of_year_type != "month_of_year",]
```

Before we go, let's look at an example that illustrates the benefits of the statistics API.
In the example below, we pull all day-of-year discharge percentiles for our site.
Keep in mind that doing so *without* the statistics API would require us to download the **entire** daily period of record for this site and hand-compute these percentiles ourselves, a time- and resource-intensive process indeed.

For demonstration, we filter to the output to the January 1 day-of-year percentiles, which include a set of percentiles commonly used on WDFN webpages (e.g., [Wisconsin water conditions](https://waterdata.usgs.gov/state/wisconsin/)).


```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

full_por_percentiles <-
  read_waterdata_stats_por(
    monitoring_location_id = site1,
    parameter_code = "00060",
    computation = c("minimum", "maximum", "median", "percentile"),
    start_date = "01-01",
    end_date = "12-31"
  )

full_por_percentiles |>
  filter(time_of_year == "01-01" & time_of_year_type == "day_of_year") |>
  arrange(percentile) |>
  select(
    time_of_year, computation, percentile, value
  )
```

After a bit of data manipulation, we can then visualize the percentiles as "ribbons" on a plot.
The final visual shows the percentile bands as progressively darker ribbons, where the minima and maxima are shown as thin dashed curves and the median values as a solid gray curve.

```{r, message=FALSE, warning=FALSE}
doy_perc_bands_plt <-
  full_por_percentiles |>
  sf::st_drop_geometry() |>
  dplyr::filter(time_of_year_type == "day_of_year") |>
  select(time_of_year, percentile, value) |>
  distinct(time_of_year, percentile, .keep_all = TRUE) |>
  mutate(time_of_year = as.Date(time_of_year, format = "%m-%d")) |>
  pivot_wider(names_from = percentile, values_from = value) |>
  ggplot(aes(x = time_of_year)) +
  geom_ribbon(aes(ymin = `95`, ymax = `100`), fill = "#292f6b") +
  geom_ribbon(aes(ymin = `90`, ymax = `95`), fill = "#5699c0") +
  geom_ribbon(aes(ymin = `75`, ymax = `90`), fill = "#aacee0") +
  geom_ribbon(aes(ymin = `25`, ymax = `75`), fill = "#e9e9e9") +
  geom_ribbon(aes(ymin = `10`, ymax = `25`), fill = "#ebd6ab") +
  geom_ribbon(aes(ymin = `5`, ymax = `10`), fill = "#dcb668") +
  geom_ribbon(aes(ymin = `0`, ymax = `5`), fill = "#8f4f1f") +
  geom_line(aes(y = `50`), linewidth = .2, color = "black") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    x = "Month–day",
    y = "Value",
    title = "Day-of-year percentile bands"
  ) +
  theme_minimal()

doy_perc_bands_plt
```

Finally, let's overlay a specific year's daily mean data onto the plot:

```{r, message=FALSE, warning=FALSE}
daily_2024 <- read_waterdata_daily(
  monitoring_location_id = site1,
  parameter_code = "00060",
  statistic_id = "00003", 
  time = "2024-01-01/2024-12-31") |>
  mutate(time_of_year = as.Date(format(time, "%m-%d"), format = "%m-%d"))

doy_perc_bands_plt +
  geom_line(data = daily_2024, aes(y = value), linewidth = 1) +
  labs(
    title = "2024 daily discharge + percentile bands"
  )
```

## Fetching monthly and annual statistics within a date range

Unlike day- or month-of-year statistics, which summarize typical seasonal conditions across many years, the `read_waterdata_stats_daterange()` function returns summaries for specific months and years within a requested date range.
Consider the example below where we fetch the average discharge value for the month of January, 2024.
Notice that the `start_date` and `end_date` arguments are given in `YYYY-MM-DD` format.

```{r, message=FALSE, warning=FALSE}
jan_daterange_mean <-
  read_waterdata_stats_daterange(
  monitoring_location_id = site1,
  parameter_code = "00060",
  computation = "arithmetic_mean",
  start_date = "2024-01-01",
  end_date = "2024-01-31"
)

jan_daterange_mean
```

Instead of `time_of_year` and `time_of_year_type` columns, this output contains `start_date`, `end_date`, and `interval_type` columns representing the daterange over which the average was calculated.
The first row shows the average January, 2025 discharge was about 219 cubic feet per second.
We again have extra rows: the second row contains the **calendar** year 2025 average and the third contains the **water** year 2025 average.

Annual statistics will be returned for any calendar/water years than intersect with the specified date range.
Consider the example below, where the `start_date` to `end_date` range is only 93 days yet happens to intersect with calendar **and** water years 2023 and 2024.
We see this reflected in output containing monthly averages for September, 2023 through January, 2024 as well as 2023 and 2024 calendar and water year averages.

```{r, message=FALSE, warning=FALSE}
multiyear_daterange_mean <-
  read_waterdata_stats_daterange(
  monitoring_location_id = site1,
  parameter_code = "00060",
  computation = "arithmetic_mean",
  start_date = "2023-09-30",
  end_date = "2024-01-01"
)

multiyear_daterange_mean
```

Before we move on, consider the following example where we create a Monthly mean statistics table similar to what you'd find in the [Water Year Summaries](https://rconnect.chs.usgs.gov/water-year-summaries-dev/?_inputs_&render_button=1&site_no_select=%2205428500%22&wateryear_select=%222024%22).

```{r, message=FALSE, warning=FALSE}
monthly_means <-
  read_waterdata_stats_daterange(
    monitoring_location_id = site1,
    parameter_code = "00060",
    computation = "arithmetic_mean"
  ) |>
  dplyr::filter(interval_type == "month") |>
  sf::st_drop_geometry()

monthly_means |>
  # filter(start_date >= "2004-10-01" & start_date < "2025-09-01") |>
  mutate(
    Month = lubridate::month(start_date, label = TRUE),
    # reorder based on WY
    Month = factor(Month, month.abb[c(10:12, 1:9)]),
    water_year = lubridate::year(as.Date(start_date) + months(3))
  ) |>
  arrange(Month) |>
  group_by(Month) |>
  summarize(
    `Mean` = round(mean(value), 0),
    `Max (WY)` = paste0(
      round(max(value), 0),
      " (",
      water_year[which.max(value)],
      ")"
    ),
    `Min (WY)` = paste0(
      round(min(value), 0),
      " (",
      water_year[which.min(value)],
      ")"
    ),
    .groups = "drop"
  ) |>
  t() |>
  as.data.frame() |>
  knitr::kable(
    col.names = NULL
  )
```



## Statistics API quirks

The `sample_count` column indicates that there were 22 observations used to compute these averages, suggesting the site's period of record is (at least) 22 years long.
We can verify this using the timeseries-metadata API endpoint, passing in the "parent" timeseries ID used to compute the mean:

```{r}
read_waterdata_ts_meta(time_series_id =  unique(jan_por_mean$parent_time_series_id))
```

From this output, we see the begin and end dates of the POR at indeed at least 22 years apart.


