---
title: "Join by closest date"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 4
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Join by closest date}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
library(knitr)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, 
                      eval = TRUE,
                      warning = FALSE, 
                      message = FALSE,
                      fig.height=7,
                      fig.width=7)


```

Increasingly there are high frequency sensor data available for water quality data. There's a need to join the sensor and discrete data by the closest time. 

This article will discuss how to do that using the `data.table` package. 

Let's look at site "01646500", and a nearby site with a real-time nitrate-plus-nitrite sensor. Our goal is to get the discharge and nitrate-plus-nitrite sensor data aligned with the discrete water quality samples.

```{r getData}
library(dataRetrieval)

site_uv <- "01646500"
site_qw <- "USGS-01646580"
pcode_uv <- "99133"
pcode_qw <- "00631"
start_date <- as.Date("2012-01-01")
end_date <- as.Date("2020-01-01")
qw_data <- readWQPqw(site_qw, pcode_qw, 
                     startDate = start_date,
                     endDate = end_date)
uv_data <- readNWISuv(siteNumbers = site_uv, 
                      parameterCd = c(pcode_uv, "00060"),
                      startDate = start_date,
                     endDate = end_date)
```


The sensor data ("uv" data) at this particular site has 2 columns of data that are important. The first task is to combine those columns. This is rather unique to this particular site and probably won't need to be done generally.

```{r trimUVdata}
library(dplyr)

uv_trim <- uv_data |> 
  select(uv_date = dateTime, 
         val1 = X_SUNA...Discontinued._99133_00000,
         val2 = X_SUNA_99133_00000,
         flow = X_00060_00000) |> 
  mutate(val_uv = if_else(is.na(val1), val2, val1)) |> 
  select(-val1, -val2)

```

```{r showuvTrim, echo=FALSE}
knitr::kable(head(uv_trim))
```

Next we'll clean up the discrete water quality "qw" data to make it easy to follow in this tutorial. 

```{r trimQWdata}
qw_trim <- qw_data |> 
  filter(ActivityTypeCode == "Sample-Routine",
         !is.na(ActivityStartDateTime)) |> 
  select(qw_date = ActivityStartDateTime,
         val_qw = ResultMeasureValue,
         det_txt = ResultDetectionConditionText)
```

```{r showqwtrim, echo=FALSE}
knitr::kable(head(qw_trim))
```

Finally, we'll use the `data.table` package to do a join to the nearest date. The code to do that is here:

```{r}
library(data.table)

setDT(qw_trim)[, join_date := qw_date]

setDT(uv_trim)[, join_date := uv_date]

closest_dt <- uv_trim[qw_trim, on = .(join_date), roll = "nearest"]
```

`closest_dt` is a `data.table` object. It similar to a data.frame, but not identical. We can convert it to a data.frame and then use `dplyr` commands. Note: the whole analysis can be done via `data.table`, but most examples in `dataRetrieval` have used `dplyr`, which is why we bring it back to data.frame. `dplyr` also has a `join_by(closest())` option, but it is more complicated because you can only specify the closeness in either the forward or backwards direction (and we want either direction).

We can calculate "delta_time" - the difference in time between the uv and qw data. We'll probably want to add a threshold that we don't join values if they are too far apart in time. In this example, if the difference is greater than 24 hours, we'll substitute `NA`.

```{r}
qw_closest <- data.frame(closest_dt) |> 
  mutate(delta_time = difftime(qw_date, uv_date, 
                               units = "hours"),
         val_uv = if_else(abs(as.numeric(delta_time)) >= 24, NA, val_uv)) |> 
  select(-join_date)

```


```{r showqwClosest, echo=FALSE}
knitr::kable(head(qw_closest))
```

Here are a few plots to show the applications of these joins:

```{r}
library(ggplot2)

ggplot(data = qw_closest) +
  geom_point(aes(x = val_uv, y = val_qw)) +
  theme_bw() +
  xlab("Sensor") +
  ylab("Discrete")

```

```{r}
ggplot(data = qw_closest) +
  geom_point(aes(x = flow, y = val_qw)) +
  theme_bw() +
  xlab("Discharge") +
  ylab("Concentration") +
  scale_x_log10() +
  scale_y_log10()

```

```{r}
ggplot() +
  geom_line(data = uv_trim,
            aes(x = uv_date, val_uv),
            color = "lightgrey") +
  geom_point(data = qw_closest,
             aes(x = qw_date, y = val_qw),
             color= "red") +
  theme_bw() +
  ggtitle("Red dots = discrete samples, grey lines = continuous sensor") +
  xlab("") +
  ylab("Concentration")

```
