---
title: "Calculating Moving Averages and Historical Flow Quantiles"
author: "Laura DeCicco"
date: "2016-10-25"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Calculating Moving Averages and Historical Flow Quantiles}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This post will show simple way to calculate moving averages, calculate historical-flow quantiles, and plot that information. The goal is to reproduce the graph at this link:
[PA Graph](http://pa.water.usgs.gov/drought/indicators/sw/images/f30_01538000.html). The motivation for this post was inspired by a USGS colleague that that is considering creating these type of plots in R. We thought this plot provided an especially fun challenge - maybe you will, too!

First we get the data using the [dataRetrieval](https://CRAN.R-project.org/package=dataRetrieval) package. The siteNumber and parameterCd could be adjusted for other sites or measured parameters. In this example, we are getting discharge (parameter code 00060) at a site in PA. 

It may be important to note that this script is a bit lazy in handling leap days. 

## Get data using dataRetrieval

```{r message=FALSE}
library(dataRetrieval)

#Retrieve daily Q
siteNumber<-c("01538000")
parameterCd <- "00060" #Discharge
dailyQ <- readNWISdv(siteNumber, parameterCd) 
dailyQ <- renameNWISColumns(dailyQ)
stationInfo <- readNWISsite(siteNumber)

```

## Calculate moving average

Next, we calculate a 30-day moving average on all of the flow data:

```{r message=FALSE}
library(dplyr)
library(zoo)

#Check for missing days, if so, add NA rows:
if(as.numeric(diff(range(dailyQ$Date))) != (nrow(dailyQ)+1)){
  fullDates <- seq(from = min(dailyQ$Date),
                   to = max(dailyQ$Date), by="1 day")
  fullDates <- data.frame(Date = fullDates, 
                          agency_cd = unique(dailyQ$agency_cd),
                          site_no = unique(dailyQ$site_no))
  dailyQ <- fullDates %>% 
    left_join(dailyQ,
              by=c("Date","agency_cd","site_no")) %>%
    arrange(Date)
}

dailyQ <- dailyQ %>%
  mutate(rollMean = rollmean(Flow, 30, fill = NA, align = "center"),
         day.of.year = as.numeric(strftime(Date, 
                                           format = "%j")))

```

## Calculate historical percentiles

We can use the `quantile` function to calculate historical percentile flows. Then use the `loess` function for smoothing. The argument `smooth.span` defines how much smoothing should be applied. To get a smooth transistion at the start of the graph, we can add include an earlier year which is not plotted at the end. 

```{r message=FALSE}
summaryQ <- dailyQ %>%
  group_by(day.of.year) %>%
  summarize(p75 = quantile(rollMean, probs = .75, na.rm = TRUE),
            p25 = quantile(rollMean, probs = .25, na.rm = TRUE),
            p10 = quantile(rollMean, probs = 0.1, na.rm = TRUE),
            p05 = quantile(rollMean, probs = 0.05, na.rm = TRUE),
            p00 = quantile(rollMean, probs = 0, na.rm = TRUE)) 

current.year <- as.numeric(strftime(Sys.Date(), format = "%Y"))

summary.0 <- summaryQ %>%
  mutate(Date = as.Date(day.of.year - 1, 
                        origin = paste0(current.year-2,"-01-01")),
         day.of.year = day.of.year - 365)
summary.1 <- summaryQ %>%
  mutate(Date = as.Date(day.of.year - 1, 
                        origin = paste0(current.year-1,"-01-01")))
summary.2 <- summaryQ %>%
  mutate(Date = as.Date(day.of.year - 1, 
                        origin = paste0(current.year,"-01-01")),
         day.of.year = day.of.year + 365)

summaryQ <- bind_rows(summary.0, summary.1, summary.2) 


smooth.span <- 0.3

summaryQ$sm.75 <- predict(loess(p75~day.of.year, data = summaryQ, span = smooth.span))
summaryQ$sm.25 <- predict(loess(p25~day.of.year, data = summaryQ, span = smooth.span))
summaryQ$sm.10 <- predict(loess(p10~day.of.year, data = summaryQ, span = smooth.span))
summaryQ$sm.05 <- predict(loess(p05~day.of.year, data = summaryQ, span = smooth.span))
summaryQ$sm.00 <- predict(loess(p00~day.of.year, data = summaryQ, span = smooth.span))

latest.years <- dailyQ %>%
  filter(Date >= as.Date(paste0(current.year-1,"-01-01"))) %>%
  mutate(day.of.year = 1:nrow(.))

#Let's just take the middle chunk:
summaryQ <- summaryQ %>% 
  filter(day.of.year %in% 1:365)  

summaryQ <- summaryQ %>% 
  bind_rows(summaryQ,
            summaryQ) %>% 
  mutate(day.of.year = 1:nrow(.) - 365)
  

```

## Plot using base R

Many of the graphical requirements defined by the USGS are difficult to achieve in `ggplot2`. Base R plotting can be used to obtain these types of graphs:

```{r fig.cap="Simple 30-day moving average daily flow plot using base R"}

title.text <- paste0(stationInfo$station_nm,"\n",
        "Provisional Data - Subject to change\n",
        "Record Start = ", min(dailyQ$Date),
        "  Number of years = ",
        as.integer(as.numeric(difftime(time1 = max(dailyQ$Date), 
        time2 = min(dailyQ$Date),
        units = "weeks"))/52.25),
        "\nDate of plot = ",Sys.Date(),
        "  Drainage Area = ",stationInfo$drain_area_va, "mi^2")

mid.month.days <- c(15, 45, 74, 105, 135, 166, 196, 227, 258, 288, 319, 349)
month.letters <- c("J","F","M","A","M","J","J","A","S","O","N","D")
start.month.days <- c(1, 32, 61, 92, 121, 152, 182, 214, 245, 274, 305, 335)
label.text <- c("Normal","Drought Watch","Drought Warning","Drought Emergency")


plot(latest.years$day.of.year, latest.years$rollMean, 
     ylim = c(1, 1000), xlim = c(1, 733),
     log="y", axes=FALSE, type="n", xaxs="i",yaxs="i",
     ylab = "30-day moving ave",
     xlab = "")
title(title.text, cex.main = 0.75)
polygon(c(summaryQ$day.of.year,rev(summaryQ$day.of.year)),
        c(summaryQ$sm.75, rev(summaryQ$sm.25)), 
        col = "darkgreen", border = FALSE)
polygon(c(summaryQ$day.of.year,rev(summaryQ$day.of.year)),
        c(summaryQ$sm.25, rev(summaryQ$sm.10)), 
        col = "yellow", border = FALSE)
polygon(c(summaryQ$day.of.year,rev(summaryQ$day.of.year)),
        c(summaryQ$sm.10, rev(summaryQ$sm.05)), 
        col = "orange", border = FALSE)
polygon(c(summaryQ$day.of.year,rev(summaryQ$day.of.year)),
        c(summaryQ$sm.05, rev(summaryQ$sm.00)), 
        col = "red", border = FALSE)
lines(latest.years$day.of.year, latest.years$rollMean, 
      lwd=2, col = "black")
abline(v = 366)
axis(2, las=1, at=c(1,100, 1000), tck = -0.02)
axis(2, at = c(seq(1,90, by = 10)), labels = NA, tck = -0.01)
axis(2, at = c(seq(100,1000, by = 100)), labels = NA, tck = -0.01)
axis(1, at =  c(mid.month.days,365+mid.month.days), 
     labels = rep(month.letters,2), 
     tick = FALSE, line = -0.5, cex.axis = 0.75)
axis(1, at = c(start.month.days, 365+start.month.days),
     labels = NA, tck = -0.02)
axis(1, at = c(182,547), labels = c(current.year-1,current.year), 
     line = .5, tick = FALSE)
legend("bottom", label.text, horiz = TRUE,
       fill = c("darkgreen","yellow","orange","red"),
       inset = c(0, 0), xpd = TRUE, bty = "n", cex = 0.75)
box()

```


## Plot using ggplot2

Finally, we can also try to create the graph using the `ggplot2` package. The following script shows a simple way to re-create the graph in `ggplot2` with no effort on imitating desired style:


```{r fig.cap="Simple 30-day moving average daily flow plot using ggplot2", alt.text = "30-day moving average daily flow plot, no effort on style", message=FALSE, warning=FALSE, fig.height=5}
library(ggplot2)

simple.plot <- ggplot(data = summaryQ, aes(x = day.of.year)) +
  geom_ribbon(aes(ymin = sm.25, ymax = sm.75, fill = "Normal")) +
  geom_ribbon(aes(ymin = sm.10, ymax = sm.25, fill = "Drought Watch")) +
  geom_ribbon(aes(ymin = sm.05, ymax = sm.10, fill = "Drought Warning")) +
  geom_ribbon(aes(ymin = sm.00, ymax = sm.05, fill = "Drought Emergency")) +
  scale_y_log10(limits = c(1,1000)) +
  geom_line(data = latest.years, aes(x=day.of.year, y=rollMean, color = "30-Day Mean"),size=2) +
  geom_vline(xintercept = 365) 

simple.plot
```

Next, we can play with various options to do a better job to imitate the style:

```{r fig.cap="Detailed 30-day moving average daily flow plot", alt.text = "30-day moving average daily flow plot", message=FALSE, warning=FALSE}

styled.plot <- simple.plot+
  scale_x_continuous(breaks = c(mid.month.days,365+mid.month.days),
                   labels = rep(month.letters,2),
                   expand = c(0, 0),
                   limits = c(0,730)) +
  annotation_logticks(sides="l") +
  expand_limits(x=0) +
  annotate(geom = "text", 
           x = c(182,547), 
           y = 1, 
           label = c(current.year-1,current.year), size = 4) +
  theme_bw() + 
  theme(axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  labs(list(title=title.text,
            y = "30-day moving ave", x="")) +
  scale_fill_manual(name="",breaks = label.text,
                    values = c("red","orange","yellow","darkgreen")) +
  scale_color_manual(name = "", values = "black") +
   theme(legend.position="bottom")

styled.plot
  
```

