---
title: "Introduction to New USGS Services"
author: Laura A. DeCicco
editor_options: 
  chunk_output_type: console
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to New USGS Services}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE, message=FALSE}
library(knitr)
library(dataRetrieval)
library(dplyr)
library(ggplot2)

options(continue = " ",
        width = 50)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.height = 4,
  fig.width = 7
)
```


As we bid adieu to the NWIS web services, we welcome a host of new web service offering, the [USGS Water Data OGC APIs](https://api.waterdata.usgs.gov/ogcapi/v0/). This is a modern access point for USGS water data. The USGS is planning to modernize **all** web services in the near future. For each of these updates, `dataRetrieval` will create a new function to access the new services. 

This document will introduce each new function (note as time goes on, we'll update this document to include additional functions). Scroll down for tips on what you'll need to do to convert your legacy scripts to use these modern functions. 

Note the timeline for the NWIS servers being shut down is currently very uncertain. We'd recommend incorporating these new functions as soon as possible to avoid future headaches.

# Monitoring Location

The `read_USGS_monitoring_location` function replaces the `readNWISsite` function.

`r dataRetrieval:::get_description("monitoring-locations")`

To access these services on a web browser, go to <https://api.waterdata.usgs.gov/ogcapi/v0/collections/monitoring-locations>.

Here is a simple example of requesting one known USGS site:

```{r}
sites_information <- read_USGS_monitoring_location(monitoring_location_id = "USGS-01491000")
```

The output includes the following information:

```{r echo=FALSE}
knitr::kable(t(sites_information))
```

Maybe that is more information than you need. You can specify which columns get returned with the "properties" argument:

```{r}
site_info <- read_USGS_monitoring_location(monitoring_location_id = "USGS-01491000",
                                           properties = c("monitoring_locations_id",
                                                          "site_type",
                                                          "drainage_area",
                                                          "monitoring_location_name"))

knitr::kable(site_info)

```

Any of those original outputs can also be inputs to the function! So let's say we want to find all the stream sites in Wisconsin:

```{r}
sites_wi <- read_USGS_monitoring_location(state_name = "Wisconsin", 
                                          site_type = "Stream")
```

The returned data includes a column "geometry" which is a collection of simple feature (sf) points. This allows for seamless integration with the `sf` package. Here are 2 quick examples of using the `sf` object in ggplot2 and leaflet:

```{r}
library(ggplot2)

ggplot(data = sites_wi) +
  geom_sf() +
  theme_minimal()
```

```{r}
library(leaflet)
leaflet_crs <- "+proj=longlat +datum=WGS84" #default leaflet crs
leaflet(data = sites_wi |> 
                sf::st_transform(crs = leaflet_crs)) |> 
    addProviderTiles("CartoDB.Positron") |> 
    addCircleMarkers(popup = ~monitoring_location_name, 
                   radius = 0.1,
                   opacity = 1)

```

# Time Series Metadata

The `read_USGS_ts_meta` function replaces the `whatNWISdata` function.

`r dataRetrieval:::get_description("time-series-metadata")`

To access these services on a web browser, go to <https://api.waterdata.usgs.gov/ogcapi/v0/collections/time-series-metadata>.


```{r}
ts_available <- read_USGS_ts_meta(monitoring_location_id = "USGS-01491000",
                                  parameter_code = c("00060", "00010"))

```


# Daily Values

The `read_USGS_daily` function replaces the `readNWISdv` function.

`r dataRetrieval:::get_description("daily")`

To access these services on a web browser, go to <https://api.waterdata.usgs.gov/ogcapi/v0/collections/daily>.



```{r}
library(dataRetrieval)

daily_modern <- read_USGS_daily(monitoring_location_id = "USGS-01491000",
                          parameter_code = c("00060", "00010"),
                          statistic_id = "00003",
                          time = c("2023-01-01", "2025-05-06"))

nrow(daily_modern)
```

The first thing you might notice is that the new service serves data in a "long" format, which means there is just a single observations per row of the data frame (see: [Pivot Data](articles/long_to_wide.html)). 

```{r}
library(ggplot2)
ggplot(data = daily_modern) +
  geom_point(aes(x = time, y = value, 
                 color = approval_status)) +
  facet_grid(parameter_code ~ ., scale = "free") +
  theme_bw()

```

## Additional new features

* Another feature of the modern returned data frame is that it has a "geometry" column which is an `sf` class and can therefore be used in `sf` processes.

* Flexible column selection:

```{r}
dv_2 <- read_USGS_daily(monitoring_location_id = "USGS-01491000",
                     parameter_code = c("00060"),
                     properties = c("monitoring_location_id",
                                    "value",
                                    "time"),
                     statistic_id = "00003",
                     time = c("2022-10-01", "2023-10-01"))
head(dv_2)
```

* When you look at the help file for the new function `?read_USGS_daily`, you'll also notice there are many more arguments. These are mostly set by default to NA. You DO NOT need to specify all of these parameters.

* Flexible data/last_modified






# API Tokens

If you find yourself running into limits, you can request an API token here: <https://api.waterdata.usgs.gov/signup/>

Then save your token in your .Renviron file like this:

```
API_USGS_PAT = "my_super_secret_token"
```

You can use `usethis::edit_r_environ()` to edit find and open your .Renviron file. You will need to restart R for that variable to be recognized. 


# Notes on dataRetrieval development

## New Features

### Style

New functions will use a "snake case", such as "read_USGS_samples". Older functions use camel case, such as "readNWISdv". The difference is the underscore between words. This should be a handy way to tell the difference between newer modern data access, and the older traditional functions. 

### Structure

Historically, we allowed users to customize their queries via the `...` argument structure. With `...`, users needed to know the exact names of query parameters before using the function. Now, the new functions will include **ALL** possible arguments that the web service APIs support. This will allow users to use tab-autocompletes (available in RStudio and other IDEs). **Users will need to understand that they are not required to specify all of these parameters. In fact, it is not advisable: the systems can get bogged down with redundant query parameters.** We expect this will be easier for users, but it might take some time to smooth out the documentation and test usability. There may be additional consequences, such as users won't be able to build up argument lists to pass into the function.  

### Dependencies

Under the hood, `dataRetrieval` changed the dependency from `httr` to `httr2`. `httr2` is the modern R package for web requests that is actively developed/maintained. As we develop functions for the modern USGS web services, we'll continue to explore updating package dependencies. Since the new services offer geospatial output, we have recommend using the `sf` package. The `whisker` package was also included to help create POST CQL2 queries. 

