---
title: "USGS Reference Lists"
editor_options: 
  chunk_output_type: console
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{USGS Reference Lists}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE, message=FALSE}
library(knitr)
library(dataRetrieval)

options(continue = " ",
        width = 50)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.height = 4,
  fig.width = 7
)
```

## USGS Reference Lists

### Agency Codes

`r dataRetrieval:::get_description("agency-codes")`

```{r}
agency_codes <- read_waterdata_metadata("agency-codes")
head(agency_codes)
```

### Altitude Datums

`r dataRetrieval:::get_description("altitude-datums")`

```{r}
altitude_datums <- read_waterdata_metadata("altitude-datums")
head(altitude_datums)
```


### Aquifer Codes

`r dataRetrieval:::get_description("aquifer-codes")`

```{r}
aquifer_codes <- read_waterdata_metadata("aquifer-codes")
head(aquifer_codes)
```

### Aquifer Types

`r dataRetrieval:::get_description("aquifer-types")`

```{r}
aquifer_types <- read_waterdata_metadata("aquifer-types")
head(aquifer_types)
```

### Coordinate Accuracy Codes

`r dataRetrieval:::get_description("coordinate-accuracy-codes")`

```{r}
coordinate_accuracy_codes <- read_waterdata_metadata("coordinate-accuracy-codes")
head(coordinate_accuracy_codes)
```

### Coordinate Datum Codes

`r dataRetrieval:::get_description("coordinate-accuracy-codes")`

```{r}
coordinate_datum_codes <- read_waterdata_metadata("coordinate-datum-codes")
head(coordinate_datum_codes)
```

### Coordinate Method Codes

`r dataRetrieval:::get_description("coordinate-method-codes")`

```{r}
coordinate_method_codes <- read_waterdata_metadata("coordinate-method-codes")
head(coordinate_method_codes)
```

### Country Identifiers

`r dataRetrieval:::get_description("countries")`

```{r}
countries <- read_waterdata_metadata("countries")
head(countries)
```

### County Identifiers

`r dataRetrieval:::get_description("counties")`

```{r}
counties <- read_waterdata_metadata("counties")
head(counties)
```

### Hydrologic Unit Codes

`r dataRetrieval:::get_description("hydrologic-unit-codes")`

```{r}
huc_codes <- read_waterdata_metadata("hydrologic-unit-codes")
head(huc_codes)
```


### Medium Codes

`r dataRetrieval:::get_description("medium-codes")`

```{r}
medium_codes <- read_waterdata_metadata("medium-codes")
head(medium_codes)
```

### Methods

`r dataRetrieval:::get_description("methods")`

```{r}
methods <- read_waterdata_metadata("methods")
head(methods)
```

### Method Categories

`r dataRetrieval:::get_description("method-categories")`

```{r}
method_categories <- read_waterdata_metadata("method-categories")
head(method_categories)
```

### Method Citations

`r dataRetrieval:::get_description("method-citations")`

```{r}
method_citations <- read_waterdata_metadata("method-citations")
head(method_citations)
```

### Citations

`r dataRetrieval:::get_description("citations")`

```{r}
citations <- read_waterdata_metadata("citations")
head(citations)
```


### National Aquifer Codes

`r dataRetrieval:::get_description("national-aquifer-codes")`

```{r}
medium_codes <- read_waterdata_metadata("medium-codes")
head(medium_codes)
```

### Parameter Codes

`r dataRetrieval:::get_description("parameter-codes")`

```{r}
parameter_codes <- read_waterdata_metadata("parameter-codes")
head(parameter_codes)
```

### Reliability Codes

`r dataRetrieval:::get_description("reliability-codes")`

```{r}
reliability_codes <- read_waterdata_metadata("reliability-codes")
head(reliability_codes)
```

### Site Types

`r dataRetrieval:::get_description("site-types")`

```{r}
site_types <- read_waterdata_metadata("site-types")
head(site_types)
```

### State Identifiers

`r dataRetrieval:::get_description("states")`

```{r}
states <- read_waterdata_metadata("states")
head(states)
```

### Statistic Codes

`r dataRetrieval:::get_description("statistic-codes")`

```{r}
statistic_codes <- read_waterdata_metadata("statistic-codes")
head(statistic_codes)
```


### Topographic Codes

`r dataRetrieval:::get_description("topographic-codes")`

```{r}
topographic_codes <- read_waterdata_metadata("topographic-codes")
head(topographic_codes)
```

### Time Zone Codes

`r dataRetrieval:::get_description("time-zone-codes")`

```{r}
time_zone_codes <- read_waterdata_metadata("time-zone-codes")
head(time_zone_codes)
```
