---
title: "dataRetrieval Status"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{dataRetrieval Status}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This page will be updated frequently with information about the status of dataRetrieval services.

Contact CompTools@usgs.gov with additional questions.

# Overview

On March 11, 2024, NWIS **discrete water quality** services were "frozen": any public data retrieval using `readNWISqw()` did not include any new data. As of dataRetrieval v2.7.17, `readNWISqw()` has been retired and replaced by `readWQPqw()`. Learn more about the change and where to find the new samples data in our [blog](https://waterdata.usgs.gov/blog/changes-to-sample-data/).

If you have additional questions about the NWIS qw data service, email CompTools@usgs.gov.

## Latest CRAN update

* dataRetrieval now uses `httr2` instead of `httr` under the hood for constructing web service calls. The `httr2` package is considered the modern replacement for `httr`, and has support going forward (unlike `httr`). Depending on how you perform the package updates, you may need to install `httr2`. 

* dataRetrieval will give a message with the requested URL each time a web service call is made. These can be hidden by wrapping dataRetrieval calls with `suppressMessages`. That being said, it is very informative to see exactly where the data is coming from. 

* The output of the "constructURL..." functions are now `httr2` requests instead of a character string. 

* The update to `httr2` will give us more flexibility to set up the eventual replacements to the NWIS web services. Over the next year, expect to see some new and major updates to USGS data access. dataRetrieval WILL stay on the cutting edge for accessing new USGS APIs. 
* WQP continues to default to the legacy system (that does not include recent USGS discrete sample data). To access the most recent USGS data from the new "WQX3" services use the function `readWQPqw` and set the argument `legacy=FALSE` or use the function `readWQPdata` and set the argument `service = "ResultWQX3"`. Why aren't the new services set as default? This is because the WQP itself still considers those services "beta", and therefore performance is not guaranteed.  

* Finally, saving the best for last! There's a new set of functions that access the new USGS "samples-data" services. This is a USGS-specific service for discrete sample data. If you are only interested in USGS discrete water quality data (as opposed to USGS AND non-USGS discrete water quality data available from Water Quality Portal), you can use the `read_USGS_samples` function.  

Read more about it here: <https://doi-usgs.github.io/dataRetrieval/articles/samples_data.html>

## Locating USGS data using the Water Quality Portal

New USGS data (post March 11, 2024) **are temporarily not accessible** on the **main** Water Quality Portal (WQP) page (www.waterqualitydata.us). Data are still being collected, but are not available on this webpage. This limited availability is expected to last a few months.

However, new USGS data **are accessible** in a pre-release (*beta*) version of the [WQP web page](https://www.waterqualitydata.us/beta/) and new [wqx3 web services](https://waterqualitydata.us/wqx3/). Data are available in the "WQX version 3.0 format" (WQX = [Water Quality Exchange](https://exchangenetwork.net/data-exchange/wqx/)) for these new "Data Profiles" (how the data is formatted by the WQP):

* Monitoring Location
* Results - Narrow
* Results - Full Physical Chemical
* Results - Basic Physical Chemical
* Sampling Activity

Guidance on how to use the new web page and web services are available in the [User Guide](https://www.waterqualitydata.us/beta/portal_userguide/) and [Web Services Guide](https://www.waterqualitydata.us/beta/webservices_documentation/). Additional profiles will continue to be added over time. 

**Disclaimer:** During the beta period, users may encounter bugs or identify issues with the implementation of the WQX 3.0 format: we welcome (and encourage!) your feedback to help improve these offerings, just send an email to WQX@epa.gov.

The current WQP data profiles (available on the main Water Quality Portal web pages and from the current web services, https://www.waterqualitydata.us) deliver data in "WQX version 2.0" (what we're referring to as the "legacy") format. These will remain available for a period of time after the rollout of version 3.0. Eventually they will be retired, but there is not yet an estimated time line.

# What to expect: dataRetrieval specific


```{r echo=FALSE}
df <- data.frame(Function = c("readWQPdata",
                              "readWQPqw",
                              "whatWQPsites",
                              "whatWQPmetrics",
                              "whatWQPsamples",
                              "whatWQPdata",
                              "readNWISqw",
                              "readWQPsummary",
                              "whatNWISdata"),
                 Status = c("Set to legacy options by default. WQX3 options available.",
                            "Set to legacy options by default. WQX3 options available.", 
                            "Set to legacy options by default. WQX3 options available.",
                            rep("Currently only available via legacy services.", 3),
                            "Retired.",
                            "Does not have accurate information for USGS data.",
                            "Does not have accurate information for qw data."))

knitr::kable(df)

```

## readWQPqw

The `readWQPqw()` function is generally advertised as a user-friendly function since it only works with a known list of sites, parameter codes or characteristic names, and start/end dates. 

As of `dataRetrieval` 2.7.17, this function will use the default WQX version 2 dataProfile, specified by the `legacy = TRUE` argument. Setting `legacy = FALSE` will return the WQX 3.0 "narrow" dataProfile. Keep in mind the 2.0 profiles will eventually be retired. For any more flexibility, users will need to use the `readWQPdata()` function.

An example of a WQX 3.0 return:

```{r}
library(dataRetrieval)
rawPcode <- readWQPqw(siteNumbers = "USGS-01594440",
                      parameterCd = "01075",
                      legacy = FALSE)
```

Compared to using the WQX 2.0 legacy results:

```{r}
rawPcode_legacy <- readWQPqw(siteNumbers = "USGS-01594440",
                      parameterCd = "01075",
                      legacy = TRUE)

```

## readWQPdata

The `readWQPdata()` function is the most flexible function to get WQP data. Currently there are 11 legacy options and 5 options that use the new WQX 3.0 profiles. Note that `readWQPdata()` does not leverage a `legacy` argument to specify which profile version the user would like returned, but instead relies on the user's specification of `service` and `dataProfile` arguments.

### WQX 3.0

There are currently three WQX 3.0 "services" available: ResultWQX, StationWQX and ActivityWQX. The "ResultWQX" service has multiple available "dataProfiles".  

| Service        | dataProfile |
| -------------- | ----------  |
| StationWQX     |             |
| ResultWQX      | fullPhysChem |
| ResultWQX      | basicPhysChem |
| ResultWQX      | narrow |
| ActivityWQX    |         |

Examples:

```{r}
data_full <- readWQPdata(siteid = "USGS-04024315",
                         characteristicName = "pH",
                         dataProfile = "fullPhysChem",
                         service = "ResultWQX3")
ncol(data_full)

data_basic <- readWQPdata(siteid = "USGS-04024315",
                          characteristicName = "pH",
                          dataProfile = "basicPhysChem",
                          service = "ResultWQX3")
ncol(data_basic)

data_narrow <- readWQPdata(siteid = "USGS-04024315",
                           characteristicName = "pH",
                           dataProfile = "narrow",
                           service = "ResultWQX3")
ncol(data_narrow)

data_sites <- readWQPdata(siteid = "USGS-04024315",
                          characteristicName = "pH",
                          service = "StationWQX3")
ncol(data_sites)

```

### WQX 2.0 - Legacy

There are 8 services available from the legacy WQP. The Station and Result legacy services can still be accessed, but users should move to StationWQX, ResultWQX, and ActivityWQX. As other former services become available in WQX 3.0, we will update these documents.

| Service        | dataProfile | WQX 3.0 service "analog" |
| -------------- | ----------  | ----------  |
| Station     |             | StationWQX |
| Result      | resultPhysChem | ResultWQX |
| Result      | biological |
| Result      | narrowResult | ResultWQX |
| Activity    | activityAll  | ActivityWQX |
| ActivityMetric |           |   |
| Project     |            |   |
| ProjectMonitoringLocationWeighting |           |   |
| ResultDetectionQuantitationLimit |         |   |
| BiologicalMetric |     |   |

Examples: 

```{r eval=FALSE}
# Data profiles: "Organization Data" (legacy)
org_data <- readWQPdata(
  statecode = "WI",
  countycode = "Dane",
  service = "Organization"
)

# Data profiles: "Project Data"  (legacy)
project_data <- readWQPdata(
  statecode = "WI",
  countycode = "Dane",
  service = "Project"
)

# Data profiles: "Project Monitoring Location Weighting Data"  (legacy)
proj_mlwd <- readWQPdata(
  statecode = "WI",
  countycode = "Dane",
  service = "ProjectMonitoringLocationWeighting"
)

# Data profiles: "Sample Results (physical/chemical metadata)"  (legacy)
samp_data <- readWQPdata(
  siteid = "USGS-04024315",
  dataProfile = "resultPhysChem",
  service = "Result"
)

# Data profiles: "Sample Results (biological metadata)"  (legacy)
samp_bio <- readWQPdata(
  siteid = "USGS-04024315",
  dataProfile = "biological",
  service = "Result"
)

# Data profiles: "Sample Results (narrow)" (legacy)
samp_narrow <- readWQPdata(
  siteid = "USGS-04024315",
  dataProfile = "narrowResult",
  service = "Result"
)

# Data profiles: "Sampling Activity"  (legacy)
samp_activity <- readWQPdata(
  siteid = "USGS-04024315",
  dataProfile = "activityAll",
  service = "Activity"
)


# Data profile: "Sampling Activity Metrics"  (legacy)
act_metrics <- readWQPdata(
  statecode = "WI",
  countycode = "Dane",
  service = "ActivityMetric"
)

# Data profile: "Result Detection Quantitation Limit Data"  (legacy)
dl_data <- readWQPdata(
  siteid = "USGS-04024315",
  service = "ResultDetectionQuantitationLimit"
)

```

## whatNWISdata


NWIS discrete water quality services are "frozen": the returned data availability will also be frozen ONLY for "qw" data_type_cd results. All other data types should not be affected. 

When the NWIS services are decommissioned (likely in 2025): there will no longer be any "qw" information provided in the output of `whatNWISdata`. Discrete water-quality availability will be available via WQP services. More information will be provided as we learn more.

Here's an example of what will change:

```{r eval=FALSE}
what_NWIS <- whatNWISdata(siteNumber = "05114000")
nrow(what_NWIS)
[1] 407
nrow(what_NWIS[what_NWIS$data_type_cd == "qw",])
[1] 381
```

So for site "05114000", there are 381 NWIS qw parameters that have been measured. Since mid-March 2024, the data availability for those 381 parameters are frozen...even if new data are collected. Eventually those 381 rows of data will not be returned, only 26 rows of data will be returned (407-381). 

New services/functions are being developed to replace the lost functionality so check back here for updated information.


