---
title: "dataRetrieval Status"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{dataRetrieval Status}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Overview

Welcome to the `dataRetrieval` status page. This page will be updated frequently with information about the status of `dataRetrieval` services.

Please contact the Computational Tools team at CompTools@usgs.gov with questions.

# New Water Data APIs

<details>
  <summary>What is an application programming interface (API)?</summary>
  When you run a `dataRetrieval` function like `read_waterdata_samples()` or `readNWISdv()`, you are actually generating a URL that contains within it specifications of the dataset in which you are interested (e.g. which monitoring locations, characteristics, pcodes, start and end dates, etc.). The format of that URL is special: it is parsed by a USGS-specific API web service, which translates the request into a database call. It then packages the response object from the database (with the data) into a format that can be sent to and then unpacked by the user. `dataRetrieval` takes a lot of the guesswork out of this process by generating the URL, sending it to the API, and wrangling the response object into a tabular dataset.   
</details>
<br>
The USGS is in the process of creating new, [publicly available APIs](https://api.waterdata.usgs.gov/) (a.k.a. "web services") to replace the existing [WaterServices](https://waterservices.usgs.gov/). `dataRetrieval` relies upon these web services to provide monitoring location information and water quantity/quality datasets. As new web services come online, `dataRetrieval` will be furnished with new functions to accommodate these changes. 

The table below shows the **existing** NWIS functions and, if applicable, their slated replacements. Note that several functions do not have direct replacements because the new services to support them do not yet exist. We will update this table as new services and functions come online. Also note that some new functions may only be available on the "develop" branch of `dataRetrieval` (`remotes::install_github("DOI-USGS/dataRetrieval", ref = "develop")`). More information on each function's specifications (complete with examples) are available in the [Introduction to New USGS Services vignette](https://doi-usgs.github.io/dataRetrieval/articles/read_waterdata_functions.html), the [package index](https://doi-usgs.github.io/dataRetrieval/reference/index.html) and/or function documentation (e.g. `?read_waterdata_daily`).
               
```{r echo=FALSE}
df <- data.frame(
  Legacy = c(
    "readNWISqw (deprecated)",
    "readNWISdv",
    "readNWISsite",
    "whatNWISsites",
    "",
    "readNWISuv",
    "readNWISrating",
    "readNWISstat",
    "readNWISmeas",
    "readNWISpeak",
    "readNWISgwl",
    "readNWISuse",
    "readNWISdata",
    "whatNWISdata",
    "readNWISpCode"
  ),
  New = c(
    "read_waterdata_samples",
    "read_waterdata_daily",
    "read_waterdata_monitoring_location",
    "read_waterdata_ts_meta",
    "read_waterdata",
    rep("", 10)
  ),
  "Available on (branch)" = c("main (CRAN)", "main (CRAN)", "main (CRAN)", "main (CRAN)", "main (CRAN)", rep("", 10))
)

knitr::kable(df, col.names = c("WaterServices (legacy) function", "Water Data (new) function", "Available on (branch name)"))

```

If you want to learn more about the new water data APIs, check out the ["What's new with WDFN APIs?" blog post](https://waterdata.usgs.gov/blog/api-whats-new-wdfn-apis/), as well as the [documentation](https://api.waterdata.usgs.gov/docs/) available on api.waterdata.usgs.gov. 

## API Keys

Do you make *a lot* of `dataRetrieval` WaterServices calls (e.g. using functions like `readNWISdv`, `readNWISuv`) per day? ...On the order of more than 50 function calls per hour? As you switch your workflows over to the new Water Data API functions, consider grabbing yourself an API key, which will bump your limit up to 1,000 requests per hour. Head to the [sign up page](https://api.waterdata.usgs.gov/signup) to get a token.

One you have your API key, add it to your `.Renviron` file like this:

```{r}
API_USGS_PAT = "[your api key]"
```

Replace [your api key] with the alphanumeric code provided by the sign-up page. That's it! `dataRetrieval` will look for an `.Renviron` file in your directories and use it for making web service calls.

# Discrete Data

In March 2024, NWIS **discrete water quality** services were "frozen": any public data retrieval using `readNWISqw()` no longer included any new data. Concurrently, the main [Water Quality Portal (WQP) API](https://www.waterqualitydata.us/) stopped serving new and updated USGS data (we will refer to this set of web services as "legacy"). Now, new and updated data are available from the [USGS Samples API](https://waterdata.usgs.gov/download-samples/#dataProfile=site) (for USGS data only) or in the [beta version](https://www.waterqualitydata.us/beta/) of the WQP (both USGS and non-USGS data).

What does this mean for water quality data users of `dataRetrieval`? Check out the sections below for more specifics.

## Samples Data

There's a new set of functions that access the USGS "samples-data" services! If you are **only** interested in USGS discrete water quality data, you can use the `read_waterdata_samples` function.  

Read more about it in the vignette, [Introducing read_waterdata_samples](https://doi-usgs.github.io/dataRetrieval/articles/samples_data.html).

## WQP

`dataRetrieval` WQP functions continue to default to the legacy system (that does not include post-March 2024 USGS discrete sample data). The new replacement services aren't currently set as the default because the WQP team still considers these services "beta", and therefore performance is not guaranteed. Users may encounter bugs or identify issues with the implementation of the new services: we welcome (and encourage!) your feedback to help improve these offerings, just send an email to WQX@epa.gov.

The table below provides a summary of the current state of WQP functions in `dataRetrieval`.

```{r echo=FALSE}
df <- data.frame(Function = c("readWQPdata",
                              "readWQPqw",
                              "whatWQPsites",
                              "whatWQPmetrics",
                              "whatWQPsamples",
                              "whatWQPdata",
                              "readWQPsummary",
                              "whatNWISdata"),
                 Status = c("Set to legacy options by default. WQX3 options available.",
                            "Set to legacy options by default. WQX3 options available.", 
                            "Set to legacy options by default. WQX3 options available.",
                            rep("Currently only available via legacy services.", 3),
                            "Does not have accurate information for USGS data.",
                            "Does not have accurate information for qw data."))

knitr::kable(df)

```

### Leveraging the beta WQP services

The beta WQP offers data in the "WQX version 3.0 format" (WQX = [Water Quality Exchange](https://exchangenetwork.net/data-exchange/wqx/)) using new "data profiles" (how the data are formatted by the WQP). There are currently three WQX 3.0 "services" available on beta: ResultWQX3, StationWQX3 and ActivityWQX3. The "ResultWQX3" service has multiple available data profiles:  

| Service        | dataProfile |
| -------------- | ----------  |
| StationWQX3     |             |
| ResultWQX3      | fullPhysChem |
| ResultWQX3      | basicPhysChem |
| ResultWQX3      | narrow |
| ActivityWQX3    |         |

**Quickstart:** To access the most recent USGS data from the new services in `dataRetrieval`, use the function `readWQPqw` and set the argument `legacy=FALSE` or use the function `readWQPdata` and set the argument `service = "ResultWQX3"`.

If you wish to leverage a specific "ResultWQX3" data profile using the beta services, your code might look something like this, using the very flexible `readWQPdata` function:

```{r, eval = FALSE}
data_full <- readWQPdata(siteid = "USGS-04024315",
                         characteristicName = "pH",
                         dataProfile = "fullPhysChem",
                         service = "ResultWQX3")
```

On the other hand, the "StationWQX3" service requires no `dataProfile` argument:

```{r, eval = FALSE}
data_station <- readWQPdata(siteid = "USGS-04024315",
                           characteristicName = "pH",
                           service = "StationWQX3")
```

Guidance on how to use the new web page and web services are available in the [User Guide](https://www.waterqualitydata.us/beta/portal_userguide/) and [Web Services Guide](https://www.waterqualitydata.us/beta/webservices_documentation/). Additional profiles will continue to be added over time. 

## NWIS "qw" Data

As of dataRetrieval v2.7.17, `readNWISqw()` has been retired and replaced by `readWQPqw()`. The `readWQPqw()` function is generally advertised as a user-friendly function since it only works with a known list of sites, parameter codes or characteristic names, and start/end dates. Learn more about the change and where to find the new samples data in the dataretrieval [Changes to NWIS QW services vignette](https://doi-usgs.github.io/dataRetrieval/articles/qwdata_changes.html) and the WDFN blog on [Improved Public Delivery of Water Quality and Field Samples](https://waterdata.usgs.gov/blog/changes-to-sample-data/).

### A note on `whatNWISdata` and "qw" data

Currently, any "qw" data summaries provided by `whatNWISdata` are very likely out of date and incorrect: do not use this function to query by `data_type_cd = "qw"`. When the NWIS services are decommissioned (likely in 2025) there will no longer be any "qw" information provided in the output of `whatNWISdata`. Discrete water-quality data will be available via WQP services and USGS Samples only. More information will be provided as we learn more.
