---
title: "Using the dataRetrieval Stats Service"
author: "David Watkins"
date: "2016-10-05"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Using the dataRetrieval Stats Service}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

# Introduction

This script utilizes the new `dataRetrieval` package access to the [USGS Statistics Web Service](https://waterservices.usgs.gov/docs/statistics/).  We will be pulling daily mean data using the daily value service in `readNWISdata`, and using the stats service data to put it in the context of the site's history.  Here we are retrieving data for July 12th in the Upper Midwest, where a major storm system had recently passed through. You can modify this script to look at other areas and dates simply by modifying the `states` and `storm.date` objects.

To run this code, we recommend having either `dataRetreival` version 2.5.13 (currently the latest release on CRAN) or version 2.6.1 (currently the latest Github release).

# Get the data

There are two separate `dataRetrieval` calls here — one to retrieve the daily discharge data, and one to retrieve the historical discharge statistics.  Both calls are inside loops to split them into smaller pieces, to accomodate web service restrictions.  The daily values service allows only single states as a filter, so we loop over the list of states. The stats service does not allow requests of more than ten sites, so the loop iterates by groups of ten site codes.  Retrieving the data can take a few tens of seconds.  Once we have both the daily value and statistics data, the two data frames are joined by site number via [dplyr's](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) `left_join` function.  We use a [pipe](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) to send the output of the join to `na.omit()` function.  Then we add a column to the final data frame to hold the color value for each station.    
  
```{r getData, warning=FALSE, message=FALSE}
# example stats service map, comparing real-time current discharge to history for each site
# reusable for other state(s)
# David Watkins June 2016


library(maps)
library(dplyr)
library(lubridate)
library(dataRetrieval)

# pick state(s) and date
states <- c("WI", "MN", "ND", "SD", "IA")
storm.date <- "2016-07-12"

# download each state individually
for (st in states) {
  stDV <- renameNWISColumns(readNWISdata(
    service = "dv",
    parameterCd = "00060",
    stateCd = st,
    startDate = storm.date,
    endDate = storm.date
  ))
  if (st != states[1]) {
    storm.data <- full_join(storm.data, stDV)
    sites <- full_join(sites, attr(stDV, "siteInfo"))
  } else {
    storm.data <- stDV
    sites <- attr(stDV, "siteInfo")
  }
}

# retrieve stats data, dealing with 10 site limit to stat service requests
reqBks <- seq(1, nrow(sites), by = 10)
statData <- data.frame()
for (i in reqBks) {
  getSites <- sites$site_no[i:(i + 9)]
  currentSites <- readNWISstat(
    siteNumbers = getSites,
    parameterCd = "00060",
    statReportType = "daily",
    statType = c("p10", "p25", "p50", "p75", "p90", "mean")
  )
  statData <- rbind(statData, currentSites)
}

statData.storm <- statData[statData$month_nu == month(storm.date) &
  statData$day_nu == day(storm.date), ]

finalJoin <- left_join(storm.data, statData.storm)
finalJoin <- left_join(finalJoin, sites)

finalJoin[, grep("_va", names(finalJoin))] <- sapply(
  finalJoin[
    ,
    grep("_va", names(finalJoin))
  ],
  function(x) as.numeric(x)
)

# remove sites without current data
finalJoin <- finalJoin[!is.na(finalJoin$Flow), ]


# classify current discharge values
finalJoin$class <- NA

finalJoin$class[finalJoin$Flow > finalJoin$p75_va] <- "navy"
finalJoin$class[finalJoin$Flow < finalJoin$p25_va] <- "red"

finalJoin$class[finalJoin$Flow > finalJoin$p25_va &
  finalJoin$Flow <= finalJoin$p50_va] <- "green"
finalJoin$class[finalJoin$Flow > finalJoin$p50_va &
  finalJoin$Flow <= finalJoin$p75_va] <- "blue"

finalJoin$class[is.na(finalJoin$class) &
  finalJoin$Flow > finalJoin$p50_va] <- "cyan"
finalJoin$class[is.na(finalJoin$class) &
  finalJoin$Flow < finalJoin$p50_va] <- "yellow"

# take a look at the columns that we will plot later:
head(finalJoin[, c("dec_lon_va", "dec_lat_va", "class")])
```

#  Make the static plot

The base map consists of two plots.  The first makes the county lines with a gray background, and the second overlays the heavier state lines.  After that we add the points for each stream gage, colored by the column we added to `finalJoin`.  In the finishing details, `grconvertXY` is a handy function that converts your inputs from a normalized (0-1) coordinate system to the actual map coordinates, which allows the legend and scale to stay in the same relative location on different maps. 

```{r plot, fig.cap="Map discharge percentiles"}
# convert states from postal codes to full names
states <- stateCdLookup(states, outputType = "fullName")
par(pty = "s")
map("county", regions = states, fill = TRUE, col = "gray87", lwd = 0.5)
map("state", regions = states, fill = FALSE, lwd = 2, add = TRUE)
points(finalJoin$dec_lon_va,
  finalJoin$dec_lat_va,
  col = finalJoin$class, pch = 19
)
title(paste("Daily discharge value percentile rank\n", storm.date), line = 1)
par(mar = c(5.1, 4.1, 4.1, 6), xpd = TRUE)

legend.colors <- c(
  "cyan", "yellow",
  "red",
  "green", "blue",
  "navy"
)
legend.names <- c(
  "Q > P50*", "Q < P50*",
  "Q < P25",
  "P25 < Q < P50", "P50 < Q < P75",
  "Q > P75"
)

legend("bottomleft",
  inset = c(0.01, .01),
  legend = legend.names,
  pch = 19, cex = 0.75, pt.cex = 1.2,
  col = legend.colors,
  ncol = 2
)
map.scale(
  ratio = FALSE, cex = 0.75,
  grconvertX(.07, "npc"),
  grconvertY(.2, "npc")
)
text("*Other percentiles not available for these sites",
  cex = 0.75,
  x = grconvertX(0.2, "npc"),
  y = grconvertY(-0.08, "npc")
)
```

# Make an interactive plot

Static maps are great for papers and presentations. When possible, interactive maps allow the reader more flexibility to examine the data. The R leaflet package makes it easy to create useful interactive maps.

```{r leaflet, fig.height=5}
library(leaflet)

finalJoin$popup <- with(finalJoin, paste(
  "<b>", station_nm,
  "</b></br>",
  "Measured Flow:", Flow,
  "ft3/s</br>",
  "25% historical:", p25_va,
  "ft3/s</br>",
  "50% historical:", p50_va,
  "ft3/s</br>",
  "75% historical:", p75_va,
  "ft3/s"
))

leafMapStat <- leaflet(data = finalJoin) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(~dec_lon_va, ~dec_lat_va,
    color = ~class, radius = 3, stroke = FALSE,
    fillOpacity = 0.8, opacity = 0.8,
    popup = ~popup
  )

leafMapStat <- addLegend(leafMapStat,
  position = "bottomleft",
  colors = legend.colors,
  labels = legend.names,
  opacity = 0.8
)

leafMapStat
```
