---
title: "dataRetrieval Status"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{dataRetrieval Status}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This page will be updated frequently with information about the status of dataRetrieval services.

# Overview

### NWIS qw services

Discrete water samples data are undergoing modernization, and NWIS services will no longer be updated with the latest data starting mid-March 2024, with a full decommission expected 6 months later. Learn more about the upcoming change and where to find the new samples data in our [blog](https://waterdata.usgs.gov/blog/changes-to-sample-data/).


If you have additional questions about the qw data service, email CompTools@usgs.gov.


### WQP: USGS Data

Limited availability of **new** USGS data: starting mid-March accessible on the Water Quality Portal. Data will still be collected but will not be publicly available. This limited availability is expected to last a few weeks.

New USGS data will be accessible again upon release of the new Data Profiles, which will be in the WQX 3.0 format. These profiles should be available starting April 2024.

These profiles are not yet available to test. When they are, we will update this page with recommendations.
 


# What to expect: dataRetrieval specific

## readNWISqw

### Current Status:

Functioning normally, includes new data. This function will be going away soon, so consider switching to `readWQPqw` or `readWQPdata`. When the NWIS qw service is eventually shut down - the `readNWISqw` will be removed from dataRetrieval. 

### March 2024

When the NWIS services are "frozen": the returned data will not include any new data. 

### Fall 2024

When the NWIS services are decommissioned (possible September 2024): the `readNWISqw` function will be removed from `dataRetrieval`.

See `vignette("qwdata_changes", package = "dataRetrieval")` for information on how to convert your workflows from `readNWISqw` to `readWQPqw`. 

## whatNWISdata

### Current Status:

Functioning normally, includes information on all available data. 

### March 2024

When the NWIS services are "frozen" (mid-March): the returned data availability will also be frozen ONLY for "qw" data_type_cd results. All other data types should not be affected. 

### Fall 2024

When the NWIS services are decommissioned (possible September 2024): there will no longer be any "qw" information provided in the output of `whatNWISdata`. Discrete water-quality availability will be available via WQP services. More information will be provided as we learn more.

Here's an example of what will change:

```{r eval=FALSE}
what_NWIS <- whatNWISdata(siteNumber = "05114000")
nrow(what_NWIS)
[1] 407
nrow(what_NWIS[what_NWIS$data_type_cd == "qw",])
[1] 381
```

So for site "05114000", there are 381 NWIS qw parameters that have been measured. Starting mid-March 2024, the data availability for those 381 parameters will be frozen...even if new data is collected. Eventually those 381 rows of data will not be returned, only 26 rows of data will be returned (407-381). 

Once the full transition is completed, we will update with additional guidance. Water Quality Portal data availability can be obtained:

```{r eval=FALSE}
what_WQP_full <- readWQPsummary(siteid = "USGS-05114000")

library(dplyr)
what_WQP <- what_WQP_full |> 
  group_by(siteid = MonitoringLocationIdentifier, CharacteristicName) |> 
  summarise(count = sum(ResultCount, na.rm = TRUE),
            start_year = min(YearSummarized, na.rm = TRUE),
            end_year = max(YearSummarized, na.rm = TRUE))

nrow(what_WQP)
[1] 203
```

You'll notice that there are fewer rows in the `what_WQP` data frame (203) than "qw" data in the `what_NWIS` (381). This is because there are sometimes more than one parameter code that are associated with a "CharacteristicName". 

## readWQPqw and readWQPdata

### Current Status:

Functioning normally, includes all available USGS data. The Water Quality Portal does not yet include new "WQX 3.0" profiles. 

### March 2024

When the NWIS services are "frozen" (mid-March): no new USGS will be available via any Water Quality Portal query. 

### Fall 2024

When the NWIS services are decommissioned (possible September 2024): Should not affect WQP functions.

### April 2024

When WQP gets new "data profiles" (April 2024): New data profiles will include ALL USGS data (legacy and new). Shortly after these profiles are released, we plan to update dataRetrieval so that these new profiles are the default. 

# Known Dependencies

### EGRET

The `readNWISSample` function will be deprecated or updated. Plan to switch to `readWQPSample`.

### HASP

The `data_available` function will need to be updated.

If you know of additional R package dependencies that will be affected by these changes, please email CompTools@usgs.gov.




