---
title: "dataRetrieval Status"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 7
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{dataRetrieval Status}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This page will be updated frequently with information about the status of dataRetrieval services.

Contact CompTools@usgs.gov with additional questions!

# Overview

As of March 11, 2024, NWIS **discrete water quality** services are "frozen": any public data retrieval will not include any new data. 

### NWIS qw services

Discrete water quality samples data are undergoing modernization, and NWIS services pulling that discrete water quality data ('qw') are no longer being updated as of March 11, 2024. A full decommission is expected approximately 6 months later. Learn more about the upcoming change and where to find the new samples data in our [blog](https://waterdata.usgs.gov/blog/changes-to-sample-data/).


If you have additional questions about the NWIS qw data service, email CompTools@usgs.gov.


### WQP: USGS Data

**New USGS data**: temporarily not accessible on the main Water Quality Portal (WQP) page (www.waterqualitydata.us). Data are still being collected, but are not publicly available. This limited availability is expected to last a few months.

**New USGS data** _are_ accessible in a pre-release (*beta*) version of the [WQP web page](https://www.waterqualitydata.us/beta/) and new [wqx3 web services](https://waterqualitydata.us/wqx3/). Data are available in the "WQX version 3.0 format" (WQX = [Water Quality Exchange](https://exchangenetwork.net/data-exchange/wqx/)) for three new "Data Profiles" (how the data is formatted by the WQP):

* Monitoring Location (a.k.a. 'Site' or 'Station')
* Narrow results
* Full Physical Chemical results
* Basic Physical Chemical

Guidance on how to use the new web page and web services are available in the [User Guide](https://www.waterqualitydata.us/beta/portal_userguide/) and [Web Services Guide](https://www.waterqualitydata.us/beta/webservices_documentation/). Additional profiles will continue to be added to this beta release in the coming weeks. 

During the beta period users may encounter bugs or identify issues with the implementation of the WQX 3.0 format: we welcome (and encourage!) your feedback to help improve these offerings, just send an email to WQX@epa.gov.

The current WQP data profiles (available on the main Water Quality Portal web pages and from the current web services, https://www.waterqualitydata.us) deliver data in "WQX version 2.0" format. These will remain available for a period of time after the rollout of version 3.0. Eventually they will be retired, but there is not yet an estimated time line.

# What to expect: dataRetrieval specific


```{r echo=FALSE}
df <- data.frame(Function = c("readWQPdata",
                              "readWQPqw",
                              "whatWQPsites",
                              "whatWQPmetrics",
                              "whatWQPsamples",
                              "whatWQPdata",
                              "readNWISqw",
                              "readWQPsummary",
                              "whatNWISdata"),
                 Status = c("Set to WQX3 full profile by default. 'legacy' option exists.",
                            "Set to WQX3 narrow profile by default. 'legacy' option exists.", 
                            "Set to WQX3 station profile by default. 'legacy' option exists.",
                            rep("Currently only available via legacy services.", 3),
                            "Deprecated: should not be continued to use.",
                            "Does not have accurate information for USGS data.",
                            "Does not have accurate information for qw data."))

knitr::kable(df)

```




## whatNWISdata


NWIS services are "frozen": the returned data availability will also be frozen ONLY for "qw" data_type_cd results. All other data types should not be affected. 

When the NWIS services are decommissioned (possible September 2024): there will no longer be any "qw" information provided in the output of `whatNWISdata`. Discrete water-quality availability will be available via WQP services. More information will be provided as we learn more.

Here's an example of what will change:

```{r eval=FALSE}
what_NWIS <- whatNWISdata(siteNumber = "05114000")
nrow(what_NWIS)
[1] 407
nrow(what_NWIS[what_NWIS$data_type_cd == "qw",])
[1] 381
```

So for site "05114000", there are 381 NWIS qw parameters that have been measured. Starting mid-March 2024, the data availability for those 381 parameters will be frozen...even if new data is collected. Eventually those 381 rows of data will not be returned, only 26 rows of data will be returned (407-381). 

New services/functions are being developed to replace the lost functionality so check back here for updated information.




# Known Dependencies

### EGRET

The `readNWISSample` function will be deprecated or updated. Plan to switch to `readWQPSample`.

### HASP

The `data_available` function will need to be updated.

If you know of additional R package dependencies that will be affected by these changes, please email CompTools@usgs.gov.




